---
title:  "Saureus methylation"
author: "Natalia Zajac"
output:
  html_document: 
    highlight: pygments
    theme: sand
    code_folding: hide
    toc: yes
    toc_depth: 4
editor_options: 
  chunk_output_type: inline
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, echo=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(rtracklayer)
library(dplyr)
library(reshape2)
library(tidyverse)
library(cowplot)
library(patchwork)
library(data.table)
library(ezRun)
library(ggrepel)

mtx = readRDS("resources/methylation_data.rds")
mtx_annot = readRDS("resources/methylation_data_EGGnogAnnotated.rds")
prot = readRDS("resources/SA6850_prXallWide_moreMetaInfo.rds")
```

# {.tabset}

## 6850 PASN vs TSB methylation per gene

```{r , echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "gene",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-4,-5)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)

mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001,]$label
PASN_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001,]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated")

p1 + p2
```

## 6850 PASN vs TSB methylation per gene by category

```{r , echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "gene",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID, feature) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-5,-6)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB, color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)

mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m4C",]$label
PASN_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m4C",]$label
TSB_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m6A",]$label
PASN_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m6A",]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m6A")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m6A")

p3 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m4C")

p4 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m4C")

p1 + p2 + p3 + p4
```


## 6850 PASN vs TSB methylation per upstream of gene

```{r , echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "upstream",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-4,-5)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)

mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001,]$label
PASN_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001,]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated")

p1 + p2
```

## 6850 PASN vs TSB methylation per upstream of gene by category

```{r , echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "upstream",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID, feature) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-5,-6)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB,color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)

mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m4C",]$label
PASN_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m4C",]$label
TSB_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m6A",]$label
PASN_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m6A",]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m6A")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m6A")

p3 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m4C")

p4 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m4C")

p1 + p2 + p3 + p4
```


## 6850 PASN vs TSB methylation per downstream of gene

```{r , echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "downstream",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-4,-5)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)

mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001,]$label
PASN_hypomethylated = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001,]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated")

p1 + p2
```


## 6850 PASN vs TSB methylation per downstream of gene by category

```{r, echo=FALSE, warning=FALSE}
mtx_sub = mtx[mtx$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "downstream",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, ID, feature) %>% summarise(n_distinct(start), mean(gene_len)) %>% mutate(prop_methylated = `n_distinct(start)`/`mean(gene_len)`) %>% dplyr::select(c(-5,-6)) %>% pivot_wider(values_from = prop_methylated, names_from = treatment) %>% filter(group == "6850") %>% replace(is.na(.), 0.0001) 
ggplot(mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, "")),aes(PASN,TSB, color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + geom_label_repel(aes(label = label), max.overlaps = 50)


mtx_sub = mtx_sub %>% mutate(label = if_else(PASN == 0.0001 | TSB == 0.0001, ID, ""))
TSB_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m4C",]$label
PASN_hypomethylated_m4C = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m4C",]$label
TSB_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$TSB == 0.0001 & mtx_sub$feature == "m6A",]$label
PASN_hypomethylated_m6A = mtx_sub[mtx_sub$label != "" & mtx_sub$PASN == 0.0001 & mtx_sub$feature == "m6A",]$label

p1 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m6A")

p2 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m6A,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m6A")

p3 = prot[prot$LocusTag %in% str_remove(PASN_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("PASN_hypomethylated_m4C")

p4 = prot[prot$LocusTag %in% str_remove(TSB_hypomethylated_m4C,"gene-"),c(28, 39, 56, 58)] %>% ggplot(aes(FDR.PASNvsTSB_givenAncestor, p.value.PASNvsTSB_givenAncestor)) + geom_point() + ylim(0,0.05) + geom_label_repel(aes(label = gene)) + ggtitle("TSB_hypomethylated_m4C")

p1 + p2 + p3 + p4
```


## 6850 PASN vs TSB methylation per EGGNOG category

```{r, echo=FALSE, warning=FALSE}
mtx_sub = mtx_annot[mtx_annot$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26,37)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "gene",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, eggNOG_OGs, feature) %>% summarise(sites = n_distinct(start)) %>% pivot_wider(values_from = sites, names_from = treatment) %>% filter(group == "6850")
ggplot(mtx_sub,aes(PASN,TSB, color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + ggtitle("eggNOG_OGs")
```


## 6850 PASN vs TSB methylation per KEGG_ko category

```{r, echo=FALSE, warning=FALSE}
mtx_sub = mtx_annot[mtx_annot$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26,44)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "gene",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, KEGG_ko, feature) %>% summarise(sites = n_distinct(start)) %>% pivot_wider(values_from = sites, names_from = treatment) %>% filter(group == "6850")
ggplot(mtx_sub,aes(PASN,TSB, color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + ggtitle("KEGG_ko")
```


## 6850 PASN vs TSB methylation per PFAM category

```{r, echo=FALSE, warning=FALSE}
mtx_sub = mtx_annot[mtx_annot$feature %in% c("m6A", "m4C"),c(1,2,3,5,9,10,24,26,53)]
mtx_sub = mtx_sub[!is.na(mtx_sub$ID),]
mtx_sub = mtx_sub[mtx_sub$category == "gene",]
mtx_sub$gene_len = mtx_sub$gene_end - mtx_sub$gene_start
mtx_sub = mtx_sub %>% group_by(treatment, group, PFAMs, feature) %>% summarise(sites = n_distinct(start)) %>% pivot_wider(values_from = sites, names_from = treatment) %>% filter(group == "6850")
ggplot(mtx_sub,aes(PASN,TSB, color = feature)) + geom_point() + scale_x_log10() + scale_y_log10() + geom_abline() + ggtitle("PFAMS")
```
